---
title: "BDA - Assignment 6"
author: "Anonymous"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# This chunk just sets echo = TRUE as default (i.e. print all code)
knitr::opts_chunk$set(echo = TRUE)
```

Import some useful libraries

```{r, warning=FALSE}
library(aaltobda)
library(rstan)
library(bayesplot)

# rstan options
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

# Import the data
data("bioassay")
```

# Generalized linear model: Bioassay with Stan

## Exercise 1

In this exercise, we model a dose-response relation for the bioassay data using the Stan probabilistic programming language. 
We assume a Gaussian prior distribution characterized by $$\begin{bmatrix}\alpha \\ \beta \end{bmatrix} \sim \text{N}(\boldsymbol {\mu}_0, \mathbf{\Sigma}_0), \,\, \text{where}\,\, \boldsymbol{\mu}_0=\begin{bmatrix}0 \\ 10 \end{bmatrix}
\,\,\text{and}\,\,\mathbf{\Sigma}_0 = \begin{bmatrix}2^2 & 10 \\ 10 & 10^2 \end{bmatrix}.$$

The model is built in Stan as follows
```{r}
code_bioassay <- "bioassay_model.stan"
writeLines(readLines(code_bioassay))
```
Now, we can prepare the data in R with
```{r}
bioassay_data <- list(N = length(bioassay$x),
                      x = bioassay$x,
                      n = bioassay$n,
                      y = bioassay$y,
                      mu = c(0, 10),
                      Sigma = matrix(data = c(4, 10, 10, 100), nrow = 2, ncol = 2))
```

Then, we can get a fit with the following R command and sample from the posterior distribution.
```{r}
max_iters <- 2000
n_chains <- 5
fit <- stan(file = 'bioassay_model.stan', data = bioassay_data, chains = n_chains, 
            iter = max_iters, warmup = floor(max_iters/2))
```

## Exercise 2
In order to check the convergence of the chains, we can use `stan::monitor` to compute $\hat{R}$ and other convergence analytics.
$\hat{R}$ is an indicator of the between- and within-chain variance of the estimates. If the between- or within-chain estimates are generally not agreeing, we will get a larger value. A larger value thus tells that we should continue with further simulations to improve our inference of the target distribution. If the value of $\hat{R}$ is less than $1.05$, it is considered safe to use the sample.

```{r}
monitor(fit)
```
As we can see, $\alpha$ (i.e. `theta[1]`) and $\beta$ (i.e. `theta[2]`) both have an $\hat{R}_\alpha=\hat{R}_\beta=1$.
Both results are below $1.05$, which means that we can accept the sample as the chains have converged.

## Exercise 3

Here, we plot the draws from $\alpha$ and $\beta$ in a scatter plot to compare the results with Figure 3.3b in BDA3.
The x- and y-axis limits are adjusted to capture the same grid as in the book figure.

```{r, fig.width=5, fig.height=5}
p <- mcmc_scatter(fit, pars = c('theta[1]', 'theta[2]'), alpha = 0.2)
(p + labs(title = "Scatter plot of alpha and beta draws",
         x = expression(alpha),
         y = expression(beta)) + stat_density_2d(color = "black", size = .5) 
  + xlim(-4, 10) + ylim(-10, 40))
```

All in all, the results look quite similar with some slight difference due to the different choice of prior (uniform prior compared to Gaussian).

## Exercise 4

This exercise has been written and compiled in Windows using R and the package `Rstan`.
I had some installation issues in the beginning since I did not follow the instructions of installing the `Rstan` from source as I had already used it in the previous assignment.
The problems disappeared when I followed the instructions.
I thought about using Aalto's jupyter when I run into problems, but I resolved those problems locally in the end.

The hardest part about the assignment was getting the hang of Stan syntax and how the model is defined.
I believe this will get easier with practice.




